Hierarchical wrinkling in a confined permeable biogel

### 2015-02-17 fig_plis/article_plis.tex
début des remaques de Seb: - vire Video 3. Numérote les Vidéos avec des labels et des refs - vire Roman de la fin d'intro - HCl fluccule, pas GDL - tighter packing of caseins - Finally if top and bottom turned off

\begin{abstract}
The slow acidification of caseins, a family of milk proteins, is well known to result in the formation of gels, e.g. yoghurt. Less known is the partial redissolution of the gel at even lower pH. Here, using caseins as pH-responsive bulding blocks and the acidulent glucono-$\delta$-lactone, we design a model biogel film that spontaneously wrinkles in cascade, provided non-adhesive confinement. Using a combination of titration, rheology, light microscopy and confocal microscopy, we demonstrate that, during continuous acidification, the gel first shrinks and then swells, inducing winkling. Taking into account both Darcy flow though the gel and Poiseuille flow in the surrounding solvent, we develop a model that predicts correctly the wrinkling wavelength. Our results should be universal for acid-induced protein gels as it is based on pH-induced charge stabilisation/destabilisation. As such it could set a benchmark to gain fundamental insights into wrinkled biological tissues, to texture food or to design surfaces for optical purposes.
\end{abstract}

@@

Yet, benchmark experiments that explore the possibility of wrinkling in confined porous soft materials immersed in a buoyancy-matched viscous medium are in line, not only from a physics point of view, but also to get fundamental insights into biologically relevant situation. Indeed, stability analysis of a film lying on a thin viscous substrate was studied only recently~\cite{Huang2002} and remains theoretical because, on a free interface, gravity dominates over viscous substrate. In a biological context, thin tissues, e.g. epithelimum or endothelium, are naturally confined and immersed in a nearly buoyant medium, e.g. lymph, blood or mucus, a practical situation where a thin viscous substrate would set the wrinkling wavelength~\cite{Fleury2015}. Moreover, most biological films are porous, and little is known on the interplay between permeability and wrinkling~\cite{Ma2004,Longley2013}.

Here, we introduce model permeable biogels that produce hierarchical wrinkling as a result of their interplay between their gelation dynamics and the confinement conditions. In slit geometry, we slowly acidify a dispersion of caseins, a family of milk proteins. As a result, not only does the casein form a gel as expected~\cite{Roefs1986,Bremer1989,Lucey1998,Dickinson2002} -- this is a well-known effect used in yoghurt making -- but the gel starts to produce wrinkling patterns in cascade when casein adhesion to the slit walls is turned off. Using a combination of rheology, light microscopy and confocal microscopy, we demonstrate that upon acidification the casein network spontaneously shrinks then swells at the micron level and wrinkles at the millimetre scale. We systematically investigate such swelling-induced wrinkling of a porous gel film sandwiched between two viscous layers, the whole system being confined between two glass plates. As shown in Figure~\ref{fig:dynamics} and in Supplementary Videos 1 and 2, transverse confinement leads to saturation of the amplitude on both the top and bottom walls followed by a cascade of secondary and ternary buckling. Such cascade buckling has already been observed by~\cite{Roman1999} in a one-dimensional macroscopic system, a sheet of steel forced to buckle between two solid plates. In addition one has to note that the primary wavelength is much smaller than the size of the system, implying that a simple buckling scenario cannot explain the observations in Figure~\ref{fig:dynamics} and that only a wrinkling instability can be invoked.

\begin{figure}
	\includegraphics{dynamics}%
	\caption{Dynamics of pattern formation in a confined film of casein gel (caseinate 4\%w, GDL 4\%w). (a) Light transmission microscopy. Successive generations of patterns are highlighted in colour to stress the absence of growth after formation. (b) 3D reconstruction from fluorescent confocal microscopy (see Supplementary Table~\ref{tab:data}c). Scale bars are \SI{1}{\milli\metre}.}%
	\label{fig:dynamics}%
\end{figure}

The main goal of the present study is thus to identify the selection mechanism of the wavelength. In the following we first describe the physical chemistry that allows us to form such patterns. Then we  model the situation and identify two possible selection mechanisms, namely porous Darcy flow or viscous Poiseuille flow. Comparing our experimental data to these two models we finally show that a combination of both Darcy and Poiseuille mechanisms accurately describes our data. This combined model pins down the dynamical origin of the constraints exerted on the gel and nicely predicts the wrinkling wavelength.
%Poiseuille~\cite{Poiseuille1842} flows of the solvent above and below the gel film for long wavelengths and a new mechanism, Darcy~\cite{Darcy1856} flows of the solvent through the gel film for short wavelengths.

@@

\begin{figure}
	\includegraphics{acidification}%
	\caption{Acid-induced protein gel properties behave non monotonically with pH. (a) pH decrease over time in a 4\%w sodium caseinate solution acidified by 4\%w GDL. The horizontal line indicates isoelectric pH of caseins. (b) Corresponding evolution of the fraction of free caseins. Symbols are direct measures upon addition of GDL. Continuous line is deduced from $x_\text{free}$ at the same pH obtained by addition of HCl 1 molar and waiting for equilibration (insert). (c) Evolution of the storage modulus $G'$ measured in a rheometer with full adhesion to rotor and stator. (d) Evolution of pore size measured by confocal microscopy in a slit geometry with either adhesion to all walls (black line) or no adhesion to top wall, i.e. allowed syn\ae{}resis and swelling (orange dashed line).}%
	\label{fig:acidification}
\end{figure}

@@

The stability of caseins dispersion strongly depend on the pH of the solution. Supplementary Video 3 shows the titration of a caseins dispersion at 1\%w by 1 molar HCl. At pH=7, the caseins are well dispersed then then flocculate between pH=5.5 to 4 and finally at pH=3 the solution become clear again. However as one increases the caseins concentration up to 4\%w, at low pH the caseins redispersion is not total, Supplementary Video 4. The state diagram of caseins as a function of the pH is related to charge stabilisation. At neutral pH caseins are charged negatively and repel each other. As the pH decreases to the isoelectric point of the caseins (pH$\approx 4.6$), casein molecules loose their negative net charge and tend to stick to each other~\cite{Roefs1986,Bremer1989,Lucey1998,Dickinson2002}. Beyond the isoelectric point caseins acquire a positive net charge and therefore partially re-dissolve. In the inset of Figure~\ref{fig:acidification}b, we quantify this effect and measure with UV absorption spectroscopy the amount of free caseins within the 4\%w dispersion as a function of pH (see Supplementary Methods). Starting at pH=6.8 all the casein molecules are well dispersed and the free casein fraction $x_\text{free}$ is 1. Upon decreasing the pH of the solution to the isolectric point the free casein fraction drops to almost 0, almost all the casein molecule are aggregated. Further decrease of the pH of the solution lead to a partial redissolution of the casein aggregates. At pH=2 the fraction of free casein is 0.35.

@@

To form a space spanning gel, flocculation needs to be prevented. We use glucono-$\delta$-lactone (GDL) a molecule  that slowly hydrolyses into an acid and thus lowers slowly, continuously and homogeneously the pH of the medium, Figure~\ref{fig:acidification}a. This GDL controlled acidification enables the formation of a bulk gel phase below pH=5.3 with $x_\text{free}\approx 0$. This process is kinetically driven by diffusion limited cluster aggregation~\cite{Bremer1989}. As shown on Figure~\ref{fig:acidification}a and b, when adding enough GDL the pH lowers beyond the isoelectric point shifting the balance between attraction and charge repulsion at the casein level, so that casein partially dissociate from the fractal network. This results in the coexistence between a dilute phase $x_\text{free}\approx 5-20\%$ and a dense glassy network, a scenario reminiscent of arrested phase separation~\cite{Cardinaux2007,lu2008gelation}.

Here the attraction between caseins is weakened enough so that we observe the coexistence between a dilute phase $x_\text{free}\approx 5-20\%$ and a dense glassy network, a scenario reminiscent of arrested phase separation. We monitor the GDL induced gelation with a rheometer in cone plate geometry. As casein strongly adsorbs to surfaces, the strain and stress are well transmitted to the casein dispersion and we can measure the elastic and viscous modulus of the casein dispersion as a function of time, Figure~\ref{fig:acidification}c. The mechanical properties of the gel are closely related to the fraction of free casein. First, the dispersion solidifies then reaches its maximum elasticity (\SI{500}{\pascal}) at the isoelectric point and finally the elastic modulus decreases down to \SI{100}{\pascal}. This finite value indicates that the sample is still a gel.

To gain further insight into the mechanical properties of the gel we turn to micro structural observations under confocal microscopy. As shown in Supplementary Figure~\ref{fig:nonmonotonic} the gel consists in a space spanning fractal network with a cut off length scale $\xi$ that corresponds to the maximum pore size~\cite{VanDijk1986}. Permeability $\alpha$ is of the order of \SI{67000}{\square\nano\metre}, see Supplementary Figure~\ref{fig:permeability}. The fractal model fits the structure at all times but its parameters vary non monotonously during GDL acidification. As shown in Figure~\ref{fig:acidification}d and Supplementary Figure~\ref{fig:nonmonotonic}, when reaching the isolectric point $\xi$ increases up to a maximum length of \SI{7}{\micro\metre} due to the solvent expulsion out of the casein network~\cite{Lucey1998} like in viscoelastic phase separation~\cite{Tanaka2000h}. This stiffens the arms of the network and is responsible for the increase of elasticity. Beyond the isoelectric point $\xi$ decreases down to \SI{4}{\micro\metre}. As caseins regain solubility, the strands of the network swell, the pores shrink and the gel weakens.

This process is reminiscent of arrested phase separation where a dilute phase coexist with a dense glassy network.

@@

Boundary conditions is the final parameter that is essential to form wrinkled patterns. To turn off the adsorption of the caseins on a surface, we coat the surface with polymer brushes (see Methods), effectively creating non adhesive boundary conditions for the gel. The evolution of $\xi$ when sticky boundaries are on reflects the inner constrains of the gel: $\xi$ increases, the stands of the network become thinner and the gel wants to retract; then $\xi$ decreases, the stands of the network swell and the gel wants to expend. When the casein adhesion to the upper boundary is turned off, the gel still not wrinkles because it remains attached to the bottom wall, but it detaches from the top wall. In this situation we can monitor $\xi$, Figure~\ref{fig:acidification}d. $\xi$ remains mostly constant but the volume occupied by the gel varies, as will be reported below.

@@

Figure~\ref{fig:sideview}a shows confocal measurements in the case where both top and bottom walls are coated, adhesion turned off, a configuration favourable to wrinkling (see also Supplementary Videos 7 and 8). Initially the dispersion is homogeneous and fills the entire slit. The black margins correspond to the glass walls. Around \SI{23}{\minute}, the gel forms and simultaneously becomes thinner and more concentrated - the fluorescence signal which is proportional to the casein concentration is brighter compared to the initial situation. This effect is the fingerprint of syn\ae{}resis: as gelation proceeds, the solvent is expelled on both the upper and lower sides. We describe this situation as a porous elastic film of thickness $h$ sandwiched between two viscous layers of respective thickness $H_1$ and $H_2$ in a cell of thickness $e=H_1+h+H_2$ (Figure~\ref{fig:sideview}b). Extracting the top and bottom surface of the gel phase allows for quantitative measurements of the relative volume (Figure~\ref{fig:sideview}c) and excess surface area of the gel film (Figure~\ref{fig:sideview}d) (see Supplementary Methods). As expected from the system design, the evolution of the volume is non-monotonous. After reaching a minimum around 35\% of its initial value at \SI{23}{\minute}, the volume increases. Shortly after,  at \SI{36}{\minute}, the planar film destabilises and starts to wrinkle in the $z$-direction creating excess area. This confirms that the instability is due to swelling. Around \SI{38}{\minute}, the amplitude of the wrinkles reach the thickness of the slit. The size of the primary pattern is now settled, Supplementary Figure~\ref{fig:growth}. However swelling continues, the gel flattens on both top and bottom walls (\SI{44}{\minute}) and finally buckles back (\SI{53}{\minute}) to form the secondary pattern (\SI{1}{\hour}21). When the excess volume saturates, the dynamics stops. The amount of GDL controls the ability of the gel to swell: at low GDL concentrations we only observe the primary pattern, at high GDL concentrations we observe the formation of higher order patterns.

Note that adhesion to side walls is necessary to prevent total syn\ae{}resis and thus allows subsequent planar stresses to develop while swelling: the only situation where we observe wrinkling is when the adhesion on both top and bottom wall is turned off. We also confirm that at any given point in time and space the thickness $h$ of the film of gel is homogeneous. The contrast in transmission microscopy (Figure~\ref{fig:dynamics}a) is therefore not due to thickness inhomogeneities but to altitude gradients. Indeed the length of the optical path along the vertical direction $z$ through the gel phase is larger where the gel makes a slope compared to a flat situation, as indicated by the yellow vertical lines in Figure~\ref{fig:sideview}b.

\begin{figure}
	\includegraphics{sideview}
	\caption{Measurements and modelling. (a) Confocal $(x,z)$ cuts showing syn\ae{}resis, swelling, wrinkling and cascade buckling. The scale bar is \SI{100}{\micro\metre} (real size ratio). (b) Schematic side view of the cell. The constant thickness gel film (orange) is surrounded by solvent. Yellow lines show the path of transmitted light through the gel phase. A region of interest is highlighted in gray. (c-d) Confocal microscopy measurement of the volume of the gel phase relative to cell volume and excess area. Crosses correspond to times in (a).}
	\label{fig:sideview}
\end{figure}

@@

The determination of the characteristic length $\lambda$ of a wrinkling pattern is an old problem. Large deformations of thin sheets is governed by F\"{o}ppl-von K\'arm\'an equations~\cite{landau1986}, which are essentially impossible to solve in analytical form. However using scaling and asymptotic arguments one can obtain the dominant wavelength of the wrinkling pattern and isolate the physically meaningful ingredients~\cite{Cerda2003}. An elastic film buckles due to excess area with respect to its boundaries. Wrinkling is a buckling hindered by a substrate: the mismatch of elastic properties between the film and the substrate selects the preferred wrinkling wavelength~\cite{Gough1940, Bijlaard1946}. $\lambda$ is then set by the competition between the flexural modulus of the film $B=E/(1-\nu^2)$ ($E$ Young modulus, $\nu$ Poisson ratio) and a transverse load $\sigma_\perp$. For small deflections of  amplitude $A$, $\lambda$ scales as~\cite{Vella2009,Kolinski2009}: 
%
\begin{equation}
\lambda \sim \left( A h^3 \frac{B}{\sigma_{\perp}} \right)^{1/4}.
\label{eq:lstar}
\end{equation}

@@

\begin{figure*}
	\includegraphics{Darcy_vs_Poiseuille}
	\caption{Comparing model predictions with measured wavelengths. Dots come from primary pattern, squares from secondary blisters. Lines are the best linear fit through the origin taking only into account the points that should be (a) in Darcy mode $H<H^*$, (b) in Poiseuille mode $H>H^*$ (c) all points. Prefactors are 0.63, 0.69 and 0.67 respectively. The dashed line is the best affine fit ($\lambda_{\rm exp}=0.52\lambda_{P}+\SI{0.33}{\milli\metre}$) to all the points in (b).}
	\label{fig:DarcyPoiseuille}
\end{figure*}

We consider two limit scenarios that leads to wrinkling patterns. In a first limit, the film of gel sits without sticking to the bottom wall of the cell, i.e. $H_1\rightarrow0$, see Supplementary Video 7. Creating a 2D blister requires flowing solvent of viscosity $\eta$ through the gel of permeability $\alpha$. Darcy law~\cite{Darcy1856} relates the pressure gradient, estimated as $\sigma_{\perp}/h$, to the volume flux per unit area $v = \frac{\alpha}{\eta}\frac{\sigma_{\perp}}{h} = \frac{dA}{dt}$. Injecting this expression in equation~(\ref{eq:lstar}) yields the wavelength of the Darcy mode as
\begin{equation}
\lambda_D \sim h^{1/2} \alpha^{1/4} \Upsilon^{1/4}
\end{equation}
where $\Upsilon = B\tau/\eta$ is a dimensionless viscoelastic factor quantifying the relative stiffness of the film and the solvent at a characteristic time $\tau$. This expression emphasises the kinetic nature of the wavelength selection through Darcy flow. However we note that the wavelength of a generation of the pattern does not evolve with time once the gel is touching both walls(see Figure~\ref{fig:dynamics}a and Supplementary Figure~\ref{fig:growth}). This arrest may be due to bending rigidity~\cite{LeGoff2014}. Here the relevant time scale $\tau$ corresponds to the times it takes for the wrinkle amplitude to saturate (see Supplementary Method for the precise definition).

The opposite limiting case is when the gel film lies in the middle of the cell ($H_1=H_2$, see Supplementary Video 8) and impermeable ($\alpha\rightarrow 0$). Here destabilisation over a wavelength $\lambda$ creates a lubrication ($H\ll\lambda$) flow in the viscous layers. By symmetry the transverse load across the gel $\sigma_{\perp}$ is also the pressure difference over the wavelength. Using a Poiseuille profile for the flow, $\frac{dA}{dt} \sim \sigma_{\perp} \frac{H^3}{\eta\lambda^2}$~\cite{Poiseuille1842}. Injecting this expression in equation~(\ref{eq:lstar}) yields the wavelength of the Poiseuille mode as
\begin{equation}
\lambda_P \sim (hH)^{1/2} \Upsilon^{1/6}
\end{equation}
where we recognise the scaling for the wrinkles of an elastic film on a thin ($H\ll\lambda$) elastic substrate~\cite{Cerda2003} where $\Upsilon$ would be the ratio between the flexural modulus of the film and that of the substrate. Indeed within the lubrication approximation or at low Reynolds numbers one can consider a viscous film as elastic with effective Young modulus $3\eta/\tau$~\cite{Biot1957,Boudaoud2001}.

A more complete derivation of the models can be found in the Supplementary Method and yields the following prefactors:
\begin{align}
\lambda_D &= 2\pi h^{1/2}\alpha^{1/4}\left(\frac{\Upsilon}{12}\right)^{1/4}\\
\lambda_P &= \pi (hH)^{1/2}\left(\frac{2}{9}\Upsilon\right)^{1/6}
\end{align}
where $1/H^3 = 1/H_1^3 + 1/H_2^3$. The crossover between the two models where $\lambda_D=\lambda_P$ is found for $H = H^* \equiv 2^{2/3} 3^{1/6} \alpha^{1/2} \Upsilon^{1/6}$.

>>>>

\begin{abstract}
The slow acidification of caseins, a family of milk proteins, is well known to result in the formation of gels, e.g. yoghurt. Less known is the partial redissolution of the gel at even lower pH. Here, using caseins as pH-responsive building blocks and the acidulent glucono-$\delta$-lactone, we design a model biogel film that spontaneously wrinkles in cascade, provided non-adhesive confinement. Using a combination of titration, rheology, light microscopy and confocal microscopy, we demonstrate that, during continuous acidification, the gel first shrinks and then swells, inducing wrinkling. Taking into account both Darcy flow through the gel and Poiseuille flow in the surrounding solvent, we develop a model that predicts correctly the wrinkling wavelength. Our results should be universal for acid-induced protein gels as they are based on pH-induced charge stabilisation/destabilisation. As such our results could set a benchmark to gain fundamental insights into wrinkled biological tissues, to texture food or to design surfaces for optical purposes.
\end{abstract}

@@

Yet, benchmark experiments that explore the possibility of wrinkling in confined porous soft materials immersed in a buoyancy-matched viscous medium are in line, not only from a physics point of view, but also to get fundamental insights into biologically relevant situations. Indeed, stability analysis of a film lying on a thin viscous substrate was performed only recently~\cite{Huang2002} and remains theoretical because, on a free interface, gravity dominates over viscous substrate. In a biological context, however, thin tissues, e.g. epithelimum or endothelium, are naturally confined and immersed in a nearly buoyant medium, e.g. lymph, blood or mucus, a practical situation where a thin viscous substrate would set the wrinkling wavelength~\cite{Fleury2015}. Moreover, most biological films are porous, and little is known on the interplay between permeability and wrinkling~\cite{Ma2004,Longley2013}.

Here, we introduce model permeable biogels that produce hierarchical wrinkling as a result of their interplay between their gelation dynamics and the confinement conditions. In slit geometry, we slowly acidify a dispersion of caseins, a family of milk proteins. As a result, not only does the casein form a gel as expected~\cite{Roefs1986,Bremer1989,Lucey1998,Dickinson2002} --this is a well-known effect used in yoghurt making-- but the gel starts to produce wrinkling patterns in cascade when casein adhesion to the slit walls is turned off, Figure~\ref{fig:dynamics} and Supplementary Videos~\ref{vid:transmitted} and \ref{vid:reconstructed}. Using a combination of titration, rheology, light microscopy and confocal microscopy, we demonstrate that upon acidification the casein network spontaneously shrinks then swells at the micron level and wrinkles at the millimetre scale. We systematically investigate such swelling-induced wrinkling of a porous gel film sandwiched between two viscous layers, the whole system being confined between two glass plates.  The main goal of the present study is to unveil the physical origin of the observed patterns and to identify the selection mechanism of the wavelength.

\begin{figure}
	\includegraphics{dynamics}%
	\caption{Dynamics of pattern formation in a confined film of casein gel (caseinate 4\%w, GDL 4\%w in water). (a) Light transmission microscopy. Successive generations of patterns are highlighted in colour to stress the absence of coarsening after formation. (b) 3D reconstruction from fluorescent confocal microscopy, see Supplementary Table~\ref{tab:data}c. Scale bars are \SI{1}{\milli\metre}.}%
	\label{fig:dynamics}%
\end{figure}

In the following we first describe the physical chemistry that allows us to form such patterns.
We then show how transverse confinement leads to saturation of the amplitude on both the top and bottom walls followed by a cascade of secondary and ternary patterns. Finally we model the situation and identify two possible selection mechanisms, namely porous Darcy flow or viscous Poiseuille flow. Comparing our experimental data to these two models we show that a combination of both Darcy and Poiseuille mechanisms accurately describes our data. This combined model pins down the dynamical origin of the constraints exerted on the gel and nicely predicts the wrinkling wavelength.
%Poiseuille~\cite{Poiseuille1842} flows of the solvent above and below the gel film for long wavelengths and a new mechanism, Darcy~\cite{Darcy1856} flows of the solvent through the gel film for short wavelengths.

@@

\begin{figure}
	\includegraphics{acidification}%
	\caption{Acid-induced protein gel properties behave non monotonically with pH. (a) pH decrease over time in a 4\%w sodium caseinate solution acidified by 4\%w GDL in water. The horizontal line indicates isoelectric pH of caseins. (b) Corresponding evolution of the fraction of free caseins. Symbols are direct measures upon addition of GDL. The continuous line is deduced from $x_\text{free}$ at the same pH obtained by addition of HCl 1 molar and waiting for equilibration (insert). (c) Evolution of the elastic modulus $G'$ measured in a rheometer with full adhesion to the cone-plane geometry. (d) Evolution of pore size measured by confocal microscopy in a slit geometry with either adhesion to all walls (black line) or no adhesion to top wall, i.e. allowed syn\ae{}resis and swelling (orange dashed line).}%
	\label{fig:acidification}
\end{figure}

@@

The stability of caseins dispersion strongly depend on the pH of the solution. Supplementary Video~\ref{vid:titration} shows the titration of a casein dispersion at 4\%w by 1 molar HCl. At pH=7, the caseins are well dispersed then they flocculate between pH=5.5 to 4 and finally at pH=3 the solution becomes clearer as caseins partially redisperse. The state diagram of caseins as a function of pH is thus related to charge stabilisation. At neutral pH caseins are charged negatively and repel each other. As the pH decreases to the isoelectric point of the caseins (pH$\approx 4.6$), casein molecules loose their negative net charge and tend to stick to each other~\cite{Roefs1986,Bremer1989,Lucey1998,Dickinson2002}. Beyond the isoelectric point caseins acquire a positive net charge and therefore partially re-dissolve. In the inset of Figure~\ref{fig:acidification}b, we quantify this effect and measure with UV absorption spectroscopy the amount of free caseins within the 4\%w dispersion as a function of pH, see Supplementary Methods. Starting at pH=6.8 all the casein molecules are well dispersed and the free casein fraction $x_\text{free}$ is 1. Upon decreasing the pH to the isolectric point the free casein fraction drops to almost 0: almost all the casein molecules are aggregated. Further decrease of the pH of the solution leads to a partial redissolution of the casein aggregates. At pH=2 the fraction of free casein is 0.35.

@@

When acidified by a strong acid like HCl, caseins flocculate and do not form a space-spanning gel. Therfore, we use glucono-$\delta$-lactone (GDL) a molecule that slowly hydrolyses into gluconic acid and thus lowers slowly, continuously and homogeneously the pH of the medium, Figure~\ref{fig:acidification}a. This GDL-controlled acidification enables the formation of a bulk gel phase below pH=5.3 with $x_\text{free}\approx 0$. This process is kinetically driven by diffusion limited cluster aggregation~\cite{Bremer1989}. As shown in Figure~\ref{fig:acidification}a and b, when adding enough GDL the pH decreases beyond the isoelectric point shifting the balance between attraction and charge repulsion at the casein level, so that caseins partially dissociate from the fractal network. This results in the coexistence between a dilute phase $x_\text{free}\approx 5-20\%$ and a dense glassy network, a scenario reminiscent of arrested phase separation~\cite{Cardinaux2007,lu2008gelation}.

We monitor the GDL-induced gelation with a rheometer in cone-plate geometry. As casein strongly adsorbs to surfaces, the strain and stress are well transmitted to the casein dispersion and we can measure the elastic and viscous moduli of the casein dispersion as a function of time, Figure~\ref{fig:acidification}c. The mechanical properties of the gel are closely related to the fraction of free casein. First, the dispersion solidifies then reaches its maximum elasticity ($\approx\SI{500}{\pascal}$) at the isoelectric point and finally the elastic modulus decreases down to $\approx\SI{100}{\pascal}$. This finite value indicates that the sample is still a gel.

To gain further insight into the mechanical properties of the gel we turn to microstructural observations under confocal microscopy. As shown in Supplementary Videos~\ref{vid:stick63} and \ref{vid:nostick63} and in Supplementary Figure~\ref{fig:nonmonotonic} the gel consists in a space spanning fractal network with a cut-off length scale $\xi$ that corresponds to the maximum pore size~\cite{VanDijk1986}. Permeability $\alpha$ is of the order of \SI{67000}{\square\nano\metre}, see Supplementary Figure~\ref{fig:permeability}. The fractal model fits the structure at all times but its parameters vary non monotonously during GDL acidification. As shown in Figure~\ref{fig:acidification}d and Supplementary Figure~\ref{fig:nonmonotonic}, when reaching the isolectric point $\xi$ increases up to a maximum length of \SI{7}{\micro\metre} due to the solvent expulsion out of the casein network~\cite{Lucey1998} like in viscoelastic phase separation~\cite{Tanaka2000h}. This corresponds to a tighter packing of the caseins within the network strands. Such stiffening of the strands explains the observed increase of elasticity. Beyond the isoelectric point caseins partially regain solubility, the packing within the strands loosens which results in weaker, thicker strands and smaller pores as $\xi$ decreases down to \SI{4}{\micro\metre}, hence the long-term decrease of the elastic modulus.

@@

Boundary conditions is the final parameter that is essential to form wrinkled patterns. In order to turn off the adsorption of the caseins on a surface, we coat the surface with polymer brushes, see Methods, effectively creating non adhesive boundary conditions for the gel. When the casein adhesion to the upper boundary is turned off, the gel still not wrinkles because it remains attached to the bottom wall, but it detaches from the top wall. In this situation we can monitor $\xi$, Figure~\ref{fig:acidification}d. $\xi$ remains mostly constant but the volume occupied by the gel varies, as will be reported below.

Finally, if the adhesion to both top and bottom walls is turned off, the film wrinkles as shown in Figure~\ref{fig:dynamics}. Such cascade buckling has already been observed in a one-dimensional macroscopic system, a sheet of steel forced to buckle between two solid plates~\cite{Roman1999}. In addition one has to note that the primary wavelength is much smaller than the size of the system, implying that a simple buckling scenario cannot explain the observations in Figure~\ref{fig:dynamics} and that only a wrinkling instability can be invoked.

@@

Figure~\ref{fig:sideview}a shows successive side views of confocal measurements in the case where both top and bottom walls are coated, adhesion turned off, a configuration that induces wrinkling, see also Supplementary Videos~\ref{vid:cutDarcy} and \ref{vid:cutPoiseuille}. Initially the dispersion is homogeneous and fills the entire slit. The black margins correspond to the glass walls. Around \SI{23}{\minute}, the gel forms and simultaneously becomes thinner and more concentrated --the fluorescence signal which is proportional to the casein concentration is brighter compared to the initial situation. This effect is the fingerprint of syn\ae{}resis: as gelation proceeds, the solvent is expelled on both the upper and lower sides. We describe this situation as a porous elastic film of thickness $h$ sandwiched between two viscous layers of respective thickness $H_1$ and $H_2$ in a cell of thickness $e=H_1+h+H_2$,  Figure~\ref{fig:sideview}b. Extracting the top and bottom surface of the gel phase allows for quantitative measurements of the relative volume, Figure~\ref{fig:sideview}c, and excess surface area of the gel film, Figure~\ref{fig:sideview}d, see Supplementary Methods. As expected from the system design, the evolution of the gel volume is non-monotonous. After reaching a minimum around 35\% of its initial value at \SI{23}{\minute}, the volume increases. Shortly after,  at \SI{36}{\minute}, the planar film destabilises and starts to wrinkle in the $z$-direction creating excess area. This confirms that the instability is due to swelling. Around \SI{38}{\minute}, the amplitude of the wrinkles reaches the thickness of the slit. The size of the primary pattern is now fixed, Supplementary Figure~\ref{fig:growth}. However swelling continues, the gel flattens on both top and bottom walls (\SI{44}{\minute}) and finally buckles back (\SI{53}{\minute}) to form the secondary pattern (\SI{1}{\hour}21). When the excess area saturates, the dynamics stop (\SI{3}{\hour}). The amount of GDL controls the ability of the gel to swell: at low GDL concentrations we only observe the primary pattern, at high GDL concentrations we observe the formation of higher order patterns.

Note that adhesion to side walls is necessary to prevent total syn\ae{}resis and thus allows subsequent planar stresses to develop while swelling: the only situation where we observe wrinkling is when the adhesion on both top and bottom wall is turned off. We also confirm that at any given point in time the thickness $h$ of the film of gel is spatially homogeneous. The contrast in transmission microscopy, Figure~\ref{fig:dynamics}a, is therefore not due to thickness inhomogeneities but to altitude gradients. Indeed the length of the optical path along the vertical direction $z$ through the gel phase is larger where the gel makes a slope compared to a flat situation, as indicated by the yellow vertical lines in Figure~\ref{fig:sideview}b.

\begin{figure}
	\includegraphics{sideview}
	\caption{Measurements and modelling. (a) Confocal $(x,z)$ cuts showing syn\ae{}resis, swelling, wrinkling and cascade buckling. The scale bar is \SI{100}{\micro\metre} (real size ratio). (b) Schematic side view of the cell. The constant thickness gel film (orange) is surrounded by solvent. Yellow lines show the path of transmitted light through the gel phase. (c-d) Confocal microscopy measurement of the volume of the gel phase relative to cell volume and excess area. Crosses correspond to times in (a).}
	\label{fig:sideview}
\end{figure}

@@

The determination of the characteristic length $\lambda$ of a wrinkling pattern is an old problem. Large deformations of thin sheets is governed by F\"{o}ppl-von K\'arm\'an equations~\cite{landau1986}, which are essentially impossible to solve in analytical form. However using scaling and asymptotic arguments one can obtain the dominant wavelength of the wrinkling pattern up to dimensionless prefactors and isolate the physically meaningful ingredients~\cite{Cerda2003}. An elastic film buckles due to excess area with respect to its boundaries. Wrinkling is a buckling hindered by a substrate: the mismatch of elastic properties between the film and the substrate selects the preferred wrinkling wavelength~\cite{Gough1940, Bijlaard1946}. $\lambda$ is then set by the competition between the flexural modulus of the film $B=E/(1-\nu^2)$ ($E$ Young modulus, $\nu$ Poisson ratio) and a transverse load $\sigma_\perp$. For small deflections of  amplitude $A$, $\lambda$ scales as~\cite{Vella2009,Kolinski2009}: 
%
\begin{equation}
\lambda \sim \left( A h^3 \frac{B}{\sigma_{\perp}} \right)^{1/4}.
\label{eq:lstar}
\end{equation}

@@

\begin{figure*}
	\includegraphics{Darcy_vs_Poiseuille}
	\caption{Comparing model predictions $\lambda_D$, $\lambda_P$ and $\lambda_{D+P}$ with measured wavelengths $\lambda_\text{exp}$. Dots come from primary pattern, squares from secondary blisters. Lines are the best linear fits through the origin taking into account only the points that should be (a) in Darcy mode $H<H^*$, (b) in Poiseuille mode $H>H^*$ (c) all points. Prefactors are 0.63, 0.69 and 0.67 respectively. The dashed line in (b) is the best affine fit ($\lambda_{\rm exp}=0.52\lambda_{P}+\SI{0.33}{\milli\metre}$) to all data points.}
	\label{fig:DarcyPoiseuille}
\end{figure*}

Therefore we consider two limit scenarios that leads to wrinkling patterns. In a first limit, the film of gel sits without sticking to the bottom wall of the cell, i.e. $H_1\rightarrow0$, see Supplementary Video~\ref{vid:cutDarcy}. Creating a two dimensional blister requires flowing solvent of viscosity $\eta$ through the gel of permeability $\alpha$. Darcy law~\cite{Darcy1856} relates the pressure gradient, $(p_2-p_1)/h$, to the volume flux per unit area $v = \frac{\alpha}{\eta}\frac{p_2-p_1}{h} = \frac{dA}{dt}$. Identifying $\sigma_{\perp}=p_2-p_1$ and injecting this expression in equation~(\ref{eq:lstar}) yields the wavelength of the Darcy mode as
\begin{equation}
\lambda_D \sim h^{1/2} \alpha^{1/4} \Upsilon^{1/4}
\end{equation}
where $\Upsilon = B\tau/\eta$ is a dimensionless viscoelastic factor quantifying the relative stiffness of the film and the solvent at a characteristic time $\tau$. This expression emphasises the kinetic nature of the wavelength selection through Darcy flow. However we note that the wavelength of a generation of the pattern does not evolve with time once the gel is touching both walls, see Figure~\ref{fig:dynamics}a and Supplementary Figure~\ref{fig:growth}. This arrest may be due to bending rigidity~\cite{LeGoff2014}. Here the relevant time scale $\tau$ corresponds to the times it takes for the wrinkle amplitude to saturate, see Supplementary Methods for the precise definition.

The opposite limiting case is when the gel film lies in the middle of the cell ($H\equiv H_1=H_2$, see Supplementary Video~\ref{vid:cutPoiseuille}) and impermeable ($\alpha\rightarrow 0$). Here destabilisation over a wavelength $\lambda$ creates a lubrication ($H\ll\lambda$) flow in the viscous layers. By symmetry the transverse load across the gel $\sigma_{\perp}=p_2-p_1$ is also the pressure difference over the wavelength. Using a Poiseuille profile for the flow, $\frac{dA}{dt} \sim \sigma_{\perp} \frac{H^3}{\eta\lambda^2}$~\cite{Poiseuille1842} and injecting this expression in equation~(\ref{eq:lstar}) yields the wavelength of the Poiseuille mode as
\begin{equation}
\lambda_P \sim (hH)^{1/2} \Upsilon^{1/6}
\end{equation}
where we recognise the scaling for the wrinkles of an elastic film on a thin ($H\ll\lambda$) elastic substrate~\cite{Cerda2003} where $\Upsilon$ would be the ratio between the flexural modulus of the film and that of the substrate. Indeed within the lubrication approximation or at low Reynolds numbers one can consider a viscous film as elastic with effective Young modulus $3\eta/\tau$~\cite{Biot1957,Boudaoud2001}.
